Datelife Reports

### 2020-04-27 paper_3-main.Rmd
ejm comments

Then, a name matching procedure is performed across `datelife`'s chronogram database which is assembled from OToL's tree repository. Compared to other existing open tree repositories OToL's metadata rich tree store is the only one that supports search, identification, and handling of chronograms in an automatised fashion. Also, the tip names in their stored trees correspond to scientific names at or below the species level. Finally, all their chronograms come from peer-reviewed published studies generated by specialists in the targeted lineages, arguably representing expert knowledge on time of lineage divergence.
<!-- The TreeBASE repository stores trees without branch lengths, so chronograms stored there do not provide relevant information on time of lineage divergence. The Dryad repository stores all types of biological data besides chronograms, and their metadata does not allow differentiating between branch lengths units.  -->
<!-- branch length units that might represent other types of time of divergence that are not proportional to absolute (geologic) time.  -->

Information from source chronograms can be summarised with a summary statistic of tree branch lengths, such as median or mean. A  much slower, but possibly more accurate Super Distance Matrix (SDM) approach for supertree reconstruction with branch lengths [@Criscuolo2006] is also implemented via the ape package [@Paradis2004].<!-- and the hierarchical clustering method of ttol??-->
The resulting summary patristic distance matrix could be clustered with classic algorithms to return a tree. However, the resulting trees are often non-ultrameric and do not reflect the source chronogram data (see datelife_examples package). Instead, we obtained a distribution of age data from the summary matrix available for nodes on a consensus tree. The Branch Length Adjuster (BLADJ) algorithm [@Webb2008] was then used to fix known node ages and then distribute nodes with unknown ages evenly over the consensus tree, minimizing age variance in the resulting chronogram.

@@

`datelife`'s code speed was tested on an Apple iMac
with one 3.4 GHz Intel Core i5 processor. <!--No multiprocessor tests were performed-->
We registered variation in computing time of query processing and search through the database relative to number of queried taxon names.
Query processing time increases roughly linearly with number of input taxon names, and
increases considerably if TNRS service is activated. Up to ten thousand names can be processed and searched in less than 30 minutes. A name search through the chronogram database with an already processed query can be performed in less than a minute, even with a very large number of taxon names (Fig. \ref{fig:runtime1}).
<!--Speed of other datelife workflows? NO, bc it refkects speed from other software mainly. But I will put a sentence about it...
Summarizing `datelife` results processing times: shows sdm speed, clustering methods speed;
generating new trees: shows bladj speed;
Adding dates processing time: reflects more mrbayes speed than anything;
get_bold_otol_tree running time: reflects bold downloading data speed.-->
<!--Speed with different number of chronograms in database? NO-->
<!--Speed of web interface and of r package in computers with different capacities? NOOOO-->
`datelife`'s code performance was evaluated with a set of unit tests designed and
implemented with the R package testthat [@RCoreTeam2018] that were run locally
--using the devtools package [@RCoreTeam2018], and on a public server --via
GitHub, using the continuous integration tool Travis CI (<https://travis-ci.org>). At
present, unit tests cover more than 50% of `datelife`'s code (<https://codecov.io/gh/phylotastic/datelife>). <!-- unit test make code robust, verifiable, debuggable, -->

@@

test improvement in synthetic otol tree after using bold sequences. Is it really better resolved? Probably not.
Do not include this analysis in paper, maybe in examples, but i think it is not worthy.
-->
In this section we demonstrate the types of outputs that can be obtained with `datelife`, using as an example the bird family Fringillidae of true finches. We performed a higher-taxon search to obtain all data on lineage divergence available in `datelife`'s database for all recognised species within the Fringillidae (`r length(dq$cleaned_names)` spp. according to the Open Tree of Life taxonomy). There are `r length(dr)` chronograms containing at least two Fringillidae species, published in `r length(unique(names(dr)))` different studies (Fig. \ref{fig:schronograms1}).
<!-- In case we want to cite the studies in here, but they are already cited in the figure caption: [Fig. \ref{fig:schronograms1}; @barker2012going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @price2014niche]. -->
Data from these source chronograms was used to generate two types of summary chronograms, median and SDM. As explained in the "Description" section, data from source chronograms was first summarised into a single distance matrix (using the median and the SDM method respectively) and then the available node ages were used as fixed ages over a consensus tree topology, to obtain a fully dated tree with the program BLADJ (Fig. \ref{fig:summaries}). Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM. In both cases, ages are generally consistent with source ages, but see next section.
<!-- (Going from a summary distance matrix Using this method is better than using clustering algorithms  -->
<!-- and hierarchical clustering method which appears to push ages back -->

Data from source chronograms was also used to date tree topologies with no branch length information and trees with branch lengths in relative substitution rates (Figs. \ref{fig:cvbladj} and \ref{fig:cvbold}). As a form of cross validation, we used tree topologies from each study and calibrated them using information from all other source chronograms. In the absence of branch length data, the ages of internal nodes were approximately recovered in almost all cases (except for studies 3, and 5; Fig. \ref{fig:cvbladj}). Maximum tree ages were only approximately recovered in one case (study 2; Fig. \ref{fig:cvbladj}).
To exemplify dating a tree with branch lengths provided by the user, we obtained DNA sequence data from the Barcode of Life Database (BOLD) and branch lengths were successfully generated for all source chronograms. However, dating with PATHd8 (using congruified calibrations) was only successful in
<!-- two studies when expanding calibrations to make them agree (studies 3 and 5), and from  -->
three cases (studies 3, 5, and 9, shown in Fig. \ref{fig:cvbold}). From these, two trees have a different sampling than the original source chronogram, mainly because DNA data for some species is absent from the BOLD. Maximum ages are quite different from source chronograms, but this might be explained also by the differences in sampling between source chronograms and BOLD trees.
More examples and code used to generate these trees can be consulted in <https://github.com/LunaSare/datelife_examples>.

>>>>

Then, a name matching procedure is performed across `datelife`'s chronogram database which is assembled from OToL's tree repository [known as the phylesystem; @mctavish2015phylesystem]. Compared to other existing open tree repositories OToL's metadata rich tree store is the only one that supports search, identification, and handling of chronograms in an automatised fashion. Also, the tip names in their stored trees correspond to scientific names at or below the species level. Finally, all their chronograms come from peer-reviewed published studies generated by specialists in the targeted lineages, arguably representing expert knowledge on time of lineage divergence.
<!-- The TreeBASE repository stores trees without branch lengths, so chronograms stored there do not provide relevant information on time of lineage divergence. The Dryad repository stores all types of biological data besides chronograms, and their metadata does not allow differentiating between branch lengths units.  -->
<!-- branch length units that might represent other types of time of divergence that are not proportional to absolute (geologic) time.  -->

Information from pruned source chronograms can then be summarised with simple summary statistics of tree branch length, such as the median or the mean. A  much slower, Super Distance Matrix (SDM) approach for supertree reconstruction with branch lengths [@Criscuolo2006] is also implemented via the ape package [@Paradis2004].<!-- and the hierarchical clustering method of ttol??-->
<!-- describe what SDM does and why clustering does not work here -->
The resulting summary patristic distance matrix could be clustered with classic algorithms to return a tree. However, the resulting trees are often non-ultrameric and do not reflect the source chronogram data (see datelife_examples package). Instead, we obtained a distribution of age data from the summary matrix available for nodes on a consensus tree. The Branch Length Adjuster (BLADJ) algorithm [@Webb2008] was then used to fix known node ages and then distribute nodes with unknown ages evenly over the consensus tree, minimizing age variance in the resulting chronogram.

@@

`datelife`'s code speed was tested on an Apple iMac
with one 3.4 GHz Intel Core i5 processor. <!--No multiprocessor tests were performed-->
We registered variation in computing time of query processing and search through the database relative to number of queried taxon names.
Query processing time increases roughly linearly with number of input taxon names, and
increases considerably if taxonomic name resolution with TNRS [@Boyle2013] is activated.
<!-- Defining tnrs here again, bc we only talk about it two times, so it easy to forget it -->
Up to ten thousand names can be processed and searched in less than 30 minutes with the most time consuming settings.
Once names have been processed as described in methods, a name search through the chronogram database can be performed in less than a minute, even with a very large number of taxon names (Fig. \ref{fig:runtime1}).
<!--Speed of other datelife workflows? NO, bc it refkects speed from other software mainly. But I will put a sentence about it...
Summarizing `datelife` results processing times: shows sdm speed, clustering methods speed;
generating new trees: shows bladj speed;
Adding dates processing time: reflects more mrbayes speed than anything;
get_bold_otol_tree running time: reflects bold downloading data speed.-->
<!--Speed with different number of chronograms in database? NO-->
<!--Speed of web interface and of r package in computers with different capacities? NOOOO-->
`datelife`'s code performance was evaluated with a set of unit tests designed and
implemented with the R package testthat [@RCoreTeam2018] that were run both locally
with the devtools package [@RCoreTeam2018], and on a public server --via
GitHub, using the continuous integration tool Travis CI (<https://travis-ci.org>). At
present, unit tests cover more than 50% of `datelife`'s code (<https://codecov.io/gh/phylotastic/datelife>). <!-- unit test make code robust, verifiable, debuggable, -->

@@

test improvement in synthetic otol tree after using bold sequences. Is it really better resolved? Probably not.
Do not include this analysis in paper, maybe in examples, but i think it is not worthy.
-->
In this section we demonstrate the types of outputs that can be obtained with `datelife`, using as an example the bird family Fringillidae of true finches. We performed a higher-taxon search to obtain data on lineage divergence available in `datelife`'s database for all recognised species within the Fringillidae (`r length(dq$cleaned_names)` species according to the Open Tree of Life taxonomy). There are `r length(dr)` chronograms containing at least two Fringillidae species, published in `r length(unique(names(dr)))` different studies (Fig. \ref{fig:schronograms1}).
<!-- In case we want to cite the studies in here, but they are already cited in the figure caption: [Fig. \ref{fig:schronograms1}; @barker2012going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @price2014niche]. -->
Data from these source chronograms was used to generate two types of summary chronograms, median and SDM. As explained in the "Description" section, data from source chronograms was first summarised into a single distance matrix (using the median and the SDM method respectively) and then the available node ages were used as fixed ages over a consensus tree topology, to obtain a fully dated tree with the program BLADJ (Fig. \ref{fig:summaries}). Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM. With both methods, ages are generally consistent with source ages, but there are some biological examples in which this is not true (see Flaws, Limitations and Prospects section).
<!-- (Going from a summary distance matrix Using this method is better than using clustering algorithms  -->
<!-- and hierarchical clustering method which appears to push ages back -->

Data from source chronograms was also used to date tree topologies with no branch length information and trees with branch lengths in relative substitution rates (Figs. \ref{fig:cvbladj} and \ref{fig:cvbold}). As a form of cross validation, we used tree topologies from each study and calibrated them using information from all other source chronograms. In the absence of branch length data, the ages of internal nodes were approximately recovered in almost all cases (except for studies 3, and 5; Fig. \ref{fig:cvbladj}).
Maximum tree ages were only approximately recovered in one case (study 2; Fig. \ref{fig:cvbladj}).
<!-- We need to think abput the root! is it comparable across source chronograms?? not all of them have the same sampling... so probably not the same root and thus not comparable. Address this -->
To exemplify dating a tree with branch lengths provided by the user, we obtained DNA sequence data using the Barcode of Life Data System [BOLD; ratnasingham2007bold] and branch lengths were successfully generated for all source chronograms. However, dating with PATHd8 [@Britton2007] using congruified calibrations, was only successful in
<!-- two studies when expanding calibrations to make them agree (studies 3 and 5), and from  -->
three cases (studies 3, 5, and 9, shown in Fig. \ref{fig:cvbold}). From these, two trees have a different sampling than the original source chronogram, mainly because DNA data for some species is absent from the BOLD. Maximum ages are quite different from source chronograms, but this might be explained also by the differences in sampling between source chronograms and BOLD trees.
More examples and code used to generate these trees were developes on an open repository that is available for consultation and reuse at <https://github.com/LunaSare/datelife_examples>.
